---
title: "Investigating risk factors associated with BMI: easySHARE data"
author: "Nicole Augustin"
output: html_document
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In this frst piece of coursework you will work with the easySHARE data. For information on data and variables, see the [easySHARE data  guide](http://www.share-project.org/fileadmin/pdf_documentation/easySHARE_Release_8.0.0_ReleaseGuide.pdf).



## Load data

Load the easySHARE data. For information on data and variables, see the [easySHARE data  guide](http://www.share-project.org/fileadmin/pdf_documentation/easySHARE_Release_8.0.0_ReleaseGuide.pdf).

```{r easyshare}
load("data/easySHARE_rel8_0_0.rda")
dat<-easySHARE_rel8_0_0

```


Load packages:
  
```{r packs, message = FALSE}
library(ggplot2)
library(tidyverse)
library(gridExtra)
library(countrycode)
```

```{r}
length(unique(dat$mergeid))
table(dat$wave,dat$int_year)
table(dat$country, dat$wave)

```
## Pre-processing 

Notice the missing values (missing codes are explained on pg. 6 of missing codes of the easySHARE guide):

```{r}

dat[dat < 0] <- NA

data_countries <- dat[dat$country %in% c(15, 16, 23, 35), ]


```

```{r}
plot(data_full$country, data_full$br010_mod)
```


```{r}
# REMOVES COLUMNS WITH ONLY NA's and first 3 columns which don't add value

data_full <- data_countries[data_countries$wave %in% c(6), ]

data_full <- data_full[ ,  !names(data_full) %in% 
    c("mergeid", "hhid", "coupleid",
    "q34_re",
   "sp009_1_mod"  ,    "sp009_2_mod"  ,"sp009_3_mod" 
       
 ,"books_age10"     , "maths_age10"  
 ,"language_age10"  , "vaccinated"  , "childhood_health",
      
  "income_pct_w1", "income_pct_w2", "income_pct_w4", "income_pct_w8", "income_pct_w7", "income_pct_w5", 
 
 "bfi10_extra_mod" , "bfi10_agree_mod" 
, "bfi10_consc_mod" , "bfi10_neuro_mod" , "bfi10_open_mod"  
,"hc002_mod"               
        
,  "numeracy_1"          
 ,"ep005_"                        
 ,"br010_mod" , "ep009_mod"      , "ep011_mod"           
, "ep026_mod"      , "ep036_mod"       ,"ep013_mod"           
 
       )]

summary(lm(bmi ~. , data = data_full))

# 
```

```{r}
# MERGE INCOME COLUMNS if looking at more than 1 wave
# MERGE COLUMNS IF WAVE SPECIFIC 
```


First missings and coding of factors. Here is an example.

```{r pre-processing}



```

```{r}
table(is.na(dat$wave), dat$country)
table(is.na(dat$language_age10), dat$wave)
```



```{r}
table(dat$bmi2, dat$int_year)
```

```{r}

table(addNA(dat$isced1997_r))

dat$educ<- factor(dat$isced1997_r,labels=c("0 - none", "1 - Primary education ", "2 - Lower secondary education",
                          "3 - (Upper) secondary education",
                          "4 - Post-secondary non-tertiary education",
                          "5 - First stage of tertiary education","6 - Second stage of tertiary education","95: still in school", "96: other"))
table(addNA(dat$educ)) # should make labels shorter

```

Next need to pick 1 - 2 waves, and up to 4 countries.

Then look at summaries. e.g. boxplots, tables, scatter plots, histograms, bar charts,..


